\chapter{Theory behind SVD}
\label{cha:svd-theory}

Let us begin by giving the formal definition of the SVD factorization,
which is in indeed a theorem (we will restrict ourselves to real
matrices, that is, to matrices whose entries belong to the field
\R{}). \\

\begin{theorem}[Singular Value Decomposition]
\label{thm:SVD}
Let $A$ be a real matrix of $n \times m$ \imply $\exists$ orthogonal matrices
$V$ ($n \times $n) and $U$  ($m \times m$), and diagonal matrix
$\Sigma$ ($m \times n$) \suchthat:

\[
A = U \Sigma \trans{V} 
\]
\\
where $\Sigma$ has the following properties: 
\begin{align*}
 & \Sigma =  diag(\sigma_1,\dots,\sigma_p), & \textds{for} p = min(n,m)  \\
 & \sigma_1 \ge \sigma_2 \ge \dots \ge \sigma_r > 0, & \textds{for} r = rank(A) \\
 & \sigma_{r+1} = \sigma_{r+2} = \dots = \sigma_p = 0 &
\end{align*}
\end{theorem}
\hfill

Note that in the \cref{thm:SVD}, we are considering diagonal
matrices on its more generic form that does not require them to be
square; the definition of diagonal matrix $M$ can simply be that any
element other than $M_{ii}$ becomes zero. \\

Before presenting the proofs, is convenient to provide more context
about the theory behind this matrix factorization (and probably also,
part of the motivation behind). \\

\input{svd-intuit.tex}

\section{The SVD proofs}
The rest of this chapter provides more theoretical background about
the SVD decomposition, in particular, it provides two different proofs of
\cref{thm:SVD}:

\begin{itemize}
  \item Algebraic proof using the Spectral Theorem.
  \item Geometric proof (implicitly using Compactness).
\end{itemize}
\hfill

Each one of those proofs is intended to bring a
different perspective, about such an important result as SVD is. The
list is not exhaustive of course, there could be many more ways of 
proving it; but hopefully the short list presented here, will give the
reader an idea about the rich theory behind this decomposition. \\

As with any mathematical theorem, proving is done based on previous
results; since this is not a text book, we can not afford the luxury
of proving every auxiliary theorem we use. However, we made an effort
for at least mentioning explicitly the theorems; pointing to
references, when possible, about their respective proofs. For some few
cases (like the Spectral Theorem), we did include the proof of the
auxiliary theorem as well.

\input{svd-proof-spec.tex}
\input{svd-proof-geo.tex}

