\section{Derivation of the serial algorithm}

The algorithm to numerically solve the SVD problem, that we chose for
this report, is essentially the one published by Berry on his PhD
thesis (\cite{berry91}). Despite of being more than 20 years old, we can
tell that it is still widely used, in particular by LSI software. A
big part of the opensource LSI implementations that we found (see
introductory chapter), refer to either the Fortran77 SVDPACK
(\cite{svdpack}), to its C incarnation SVDPACKC (\cite{svdpackc}), or
to its even newer skin SVDLIBC (\cite{svdlibc}). All of them are
essentially the same algorithm that Berry published in his PhD
thesis. \\

We will proceed to derive the algorith in the next sub-sections. A
cautionary warning about the level of detail presented is appropriate:
although we would like to offer the same level of detail and
technicality than the theory \cref{cha:svd-theory}, time constraints
for the delivery of this report forced us just to omit the theorems
and its proofs. A pending task, for this project to evolve into a full
Msc. thesis, would be to achieve the same level of formality than
\cref{cha:svd-theory}, indeed. Such exercise is actually required, if
one pretends to offer an innovation to the problem of efficiently
compute the SVD for LSI problem. Let us consider it a pending task
then, for the time being. \\

Since we established the equivalence of the SVD problem, to that of the
eigenproblem for gramian matrix $\trans{A}A$, is important that we
keep in mind such alternate formulation the next sections to come. From now
on, to the end of this section, our goal will be to find the
eigenvalues and eigenvectors of a symmetric matrix; assuming that the
calculations are to be performed on a computer with finite
precision. We will find with simple methods, and evolve them until we
reach the level of sophistication that we require for a practical SVD
algorithm. \\

\input{svd-lanczos-serial-power.tex}
\input{svd-lanczos-serial-ritz.tex}
\input{svd-lanczos-serial-step.tex}
\input{svd-lanczos-serial-alg.tex}



