\chapter{Lanczos SVD Algorithm}

Although quite useful for understanding the SVD factorization, the
implicit algorithms mentioned in the \cref{cha:svd-theory} can not be
used directly to calculate SVD in practice. This is in part due the
intrinsic errors associated with using a finite-precision device like
a computer (while the theorems we proved the existence of SVD, assumed
infinite precision). There is a whole area of Mathematics, called
Numerical Analysis, which targets the proper translation of
theoretical algorithms into numerical ones; which can produce accurate
results on a computer. A quite notable subarea, Numerical Linear
Algebra, has received special attention over decades of research. The
previous chapter mentions some of the main aspects to consider, with
this regard. \\

Therefore, in order to calculate SVD in practice, more specialized
theorems that consider the limitations of computers are required; such
theorems allow one to create particular algorithms, which possess the
desired qualities: not only accuracy but high performance as well.
This is specially true in today's world, where the scale of the
matrices to analyze by far exceeds those used in the past. The
particular case of Latent Semantic Indexing (LSI), the particular
application we chose to restrict our study of SVD, illustrates very
well this trend in the change of scale: original papers about LSI used
a few thousands of documents, while today applications can easily
reach millions, hundreds of millions or even more (like the Oracle
product mentioned on the introductory chapter, which considers as
documents ``posts'' in social networks). \\

In this chapter we will describe the serial algorithm that is most
commonly used, for solving the SVD factorization that comes from the
LSI problem. A particular characteristic of this algorithm, at least
in the form presented here, is that is implicitly assumed that the
matrix and auxiliary data fit in the computer memory; therefore, it
will have certain limitations in terms the size of the matrix to
factorize (at least in commodity hardware, which is our focus).  The
\cref{cha:svd-dist} describes another distributed algorithm, which does allow
to spread the matrix in a cluster of commodity computers. \\

Although the algorithm presented in this chapter is essentially
serial, some of the routines it uses can accept parallelism and take
advantage of either multi-core computers with shared memory; or
vectorial processors. We mention those details in the last section as
well. 

\input{svd-lanczos-eigen.tex}
\input{svd-lanczos-serial.tex}
\input{svd-lanczos-para.tex}
